Anomaly detection with kernel density estimation on manifolds

### 2022-06-06 paper/paper.Rmd
Add kde reference

In general, given a multivariate random sample $\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel density estimate of $f$ at point $\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d$ is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
\end{equation} 
where $\pmb{H}_i$ is called the bandwidth matrix which is a $d\times d$ symmetric positive definite matrix variable at each $i$, $K(\pmb{y})$ is the kernel function, and $K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})$.
When using the standard gaussian kernel, $K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})$, we have
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
\end{equation} 
with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.

The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{y})$. We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing. 
@Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*. However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
Further, the estimator in Equation \@ref(eq:kde) is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.

>>>>

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data.

In general, given a multivariate random sample $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel density estimate of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
\end{equation} 
where $\pmb{H}_i$ is called the bandwidth matrix which is a $d\times d$ symmetric positive definite matrix variable at each $i$, $K(\pmb{x})$ is the kernel function, and $K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{x})$.
When using the standard gaussian kernel, $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$, we have
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation} 
with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.

The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{x})$. We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing. 
@Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*. However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
Further, the estimator in Equation \@ref(eq:kde) is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.

### 2022-06-06 paper/paper.tex
Add kde reference

In general, given a multivariate random sample \(\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d\), \(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel density estimate of \(f\) at point \(\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d\) is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
\end{equation}
where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{y})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})\).
When using the standard gaussian kernel, \(K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})\), we have
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
\end{equation}
with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).

The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{y})\). We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.
\textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}. However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
Further, the estimator in Equation \eqref{eq:kde} is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.

>>>>

In general, given a multivariate random sample \(\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d\), \(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel density estimate of \(f\) at point \(\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d\) is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
\end{equation}
where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{x})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{x})\).
When using the standard gaussian kernel, \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\), we have
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation}
with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).

The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{x})\). We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.
\textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}. However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
Further, the estimator in Equation \eqref{eq:kde} is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.
